---
title: "R Notebook"
output: html_notebook
---


# pick up some movies to make data smaller for testing

```{r}
train_small <- validation %>% 
  group_by(movieId) %>%
  filter(n() >= 50) %>% ungroup() %>% 
  group_by(userId) %>%
  filter(n() >= 50) %>% ungroup()

y <- train_small %>% 
  select(userId, movieId, rating) %>%
  spread(movieId, rating) %>%
  as.matrix()

rownames(y)<- y[,1]
y <- y[,-1]

movie_titles <- movielens %>% 
  select(movieId, title) %>%
  distinct()

colnames(y) <- with(movie_titles, title[match(colnames(y), movieId)])
#y_test[is.na(y_test)] <- 0
#sds <- colSds(x)
#qplot(sds, bins = 4, color = I("black"))

```


filtering to test with PCA, e.g users with rating 5
```{r}
mymovie <- train_small %>% filter(rating<2) %>% group_by(genres,movieId,title) %>% summarise(count=n()) %>%  arrange(desc(count)) %>% .$title

train_small %>% filter(rating<2) %>% group_by(genres,movieId,title) %>% summarise(count=n()) %>%  arrange(desc(count))
```

```{r}
edx %>% group_by(movieId,title, rating) %>% filter(title=='Seven (a.k.a. Se7en) (1995)') %>% summarise(n())
```


create a vector of dumb comedy
```{r}
q <- matrix(c(1 , 1, 1, -1, -1), ncol=1)
rownames(q) <- c("Ace Ventura: Pet Detective (1994)", "Ace Ventura: When Nature Calls (1995)")
```



# you can select a few columns or all columns for your PCA.

# you can use the code below to filter specific movies and find the index
```{r}
mymovie[which(str_detect(mymovie,'Ace Ventura'))]

```


```{r}
cor(y2)
```


```{r}
m_1 <- "American Pie (1999)"
m_2 <- "American Pie 2 (2001)"
qplot(y2[ ,m_1], y2[,m_2], xlab = m_1, ylab = m_2)
```


m_1 <- "Godfather, The"
m_2 <- "Godfather: Part II, The"
qplot(y[ ,m_1], y[,m_2], xlab = m_1, ylab = m_2)


```{r}
#y2 <- y[,mymovie[c(1:30,41,79,257,302,754,901,321,1572,1819,2139)]]

y2 <- y[,1:50]
```

# create residual
```{r}
y2 <- sweep(y2, 2, colMeans(y2, na.rm=TRUE))
y2 <- sweep(y2, 1, rowMeans(y2, na.rm=TRUE))
y2[is.na(y2)] <- 0
```


check residual
```{r}

library(matrixStats)
avgResidual <- rowMeans(y2)
qplot(avgResidual, bins = 2500, color = I("black"))

#sds
```

# data wrangling through matrix

I'm trying to figure out how to filter this data since at total level they are offsetting each other.

```{r}
myval <- y2[,'Pulp Fiction (1994)']

overallRating <- edx %>% group_by(movieId,title) %>% filter(title=='Pulp Fiction (1994)') %>% summarise(mean(rating), n())

#dim(as.matrix(myval))
userA <- which(myval>0 & myval<2)

myval[5]  # same result as below

y2['125',216] # same result as above

# to find col index for Thinner 1996 do the following 
which(colnames(y)=='Pulp Fiction (1994)')

```




#PCA
```{r}
y2[is.na(y2)] <- 0
pca <- prcomp(y2)

dim(pca$rotation)   # movie effect
dim(pca$x)  # user effect

plot(pca$sdev)
```


```{r}
b <- data.frame(pca$x, name = rownames(y2))
```


```{r}
library(ggrepel)

pcs <- data.frame(pca$rotation, name = colnames(y2))

highlight <- filter(pcs, PC1 < -0.1 | PC1 > 0.1 | PC2 < -0.075 | PC2 > 0.1)

pcs %>%  ggplot(aes(PC1, PC2)) + geom_point() + 
  geom_text_repel(aes(PC1, PC2, label=name),
                  data = highlight, size = 3)
```

```{r}
pcs %>% select(name, PC1) %>% arrange(PC1) %>% slice(1:15)
pcs %>% select(name, PC1) %>% arrange(desc(PC1)) %>% slice(1:15)
```

#clustering
```{r}
top <- edx %>%
  group_by(movieId) %>%
  summarize(n=n(), title = first(title)) %>%
  top_n(50, n) %>%
  .$movieId

x <- edx %>% 
  filter(movieId %in% top) %>%
  group_by(userId) %>%
  filter(n() >= 25) %>%
  ungroup() %>% 
  select(title, userId, rating) %>%
  spread(userId, rating)

row_names <- str_remove(x$title, ": Episode") %>% str_trunc(20)
x <- x[,-1] %>% as.matrix()
x <- sweep(x, 2, colMeans(x, na.rm = TRUE))
x <- sweep(x, 1, rowMeans(x, na.rm = TRUE))
rownames(x) <- row_names
```

```{r}
d <- dist(x)
h <- hclust(d)
plot(h, cex = 0.75)

groups <- cutree(h, k = 10)
split(names(groups), groups)
```

```{r}
h_2 <- dist(t(x)) %>% hclust()
```


```{r}
x_0 <- x
x_0[is.na(x_0)] <- 0

k <- kmeans(x_0, centers = 10, nstart = 25)
groups <- k$cluster
split(names(groups), groups)
```




#recommender app

# SPLIT DATA 
```{r}
set.seed(1)
library(caret)
index <- createDataPartition(edx$rating,times =1,p=0.8,list=FALSE)

train <- edx[index,]
test <- edx[-index,]

test <- test %>% 
  semi_join(train, by = "movieId") %>%
  semi_join(train, by = "userId")

dim(train)
dim(test)

```

# create y matrix 

```{r}
train_small <- train %>% 
  group_by(movieId) %>%
  filter(n() >= 50) %>% ungroup() %>% 
  group_by(userId) %>%
  filter(n() >= 50) %>% ungroup()

y <- train_small %>% 
  select(userId, movieId, rating) %>%
  spread(movieId, rating) %>%
  as.matrix()

rownames(y)<- y[,1]
y <- y[,-1]

movie_titles <- train %>% 
  select(movieId, title) %>%
  distinct()

colnames(y) <- with(movie_titles, title[match(colnames(y), movieId)])
#y_test[is.na(y_test)] <- 0
#sds <- colSds(x)
#qplot(sds, bins = 4, color = I("black"))

```


# put the initial matrix to a second matrix called y2

```{r}

y2 <- as(y, "realRatingMatrix")
rec <- Recommender(y2, method = "POPULAR")
```


# predict the outcome for users in row 8 to 10 
```{r}
pre <- predict(rec, y2[8:10], type="ratings")
pre
as(pre, "matrix")[,1:10]

```


# there are a few ways you can convert results to list, data frame etc.... see below
```{r}
#as(pre,"list")
matrix <- as(pre,"matrix")
df <- as(pre,"data.frame")

df %>% filter(user==19 & str_detect(item,'Wizard of Oz')) %>% arrange(item)

#print(rec)
##names(getModel(rec))
#getModel(rec)$nn

```

# check the results against validation data and see how accurate they are? NOT THAT MUCH DIFFERENT FROM NORMAL BIAS FROM WHAT I CAN SEE SO FAR!!!!
```{r}

validation %>% filter(userId==19) %>% arrange(title)

```


# you can change column names to tidy up..... ISSUE!!!! different type for userid (factor vs integet)    >>>>>> can't convert due to error mesages.... check later!!!
```{r}
colnames(df) <- c('userId','title','rating')   # change the column names
head(df)

```


# FINALLLYYY >>>>>>> how to query results
```{r}
as(pre,"matrix")[,1:10]     # all users for movies 1 to 10
as(pre,"matrix")[3,2]         # users at row 3 and column 2

as.integer(as(pre,"matrix")[3,2] )        # rounding to integer numbers

```

# and yeah the FUCKING!!! LIST ..... here we go!!!! arggg finally
```{r}
mylist <- as(pre,"list")

# user 1 at index 3
mylist[[1]][3]

```


# convert to data frame for user 1
```{r}
u1 <- as.data.frame(mylist[[1]])
#attributes(u1)
#class(u1)

# add column name
u1$id <- row.names(u1)

head(u1)

colnames(u1) <- c('Rating','id')
head(u1)
```

# don't know what this thing do.... check later if required!!! but for now.... dont worry!!!
```{r}
image(pre,main="Raw Ratings")
image(pre,main="Normalized Ratings")
```

